Clustering and Parameter Estimation for the Popularity Adjusted Block Model

### 2020-07-22 summary.Rmd
moving stuff from rmarkdown to separate simulation scripts

```{r}
import::from(magrittr, `%>%`)
import::from(foreach, foreach, `%do%`)
library(plot.matrix)
library(mclust)
library(ggplot2)

theme_set(theme_bw())

source('http://pages.iu.edu/~mtrosset/Courses/675/stress.r')

doMC::registerDoMC(parallel::detectCores())

embedding <- function(A, p = NULL, q = NULL,
                      eps = 1e-6) {
  eigen.A <- eigen(A, symmetric = TRUE)
  if (is.null(p) | is.null(q)) {
    keep <- (abs(eigen.A$values) > eps)
  } else {
    keep <- c(seq(p), seq(n, n - q + 1))
  }

  U <- eigen.A$vectors[, keep]
  S <- diag(sqrt(abs(eigen.A$values[keep])))
  return(U %*% S)
}

draw.graph <- function(P) {
  A <- apply(P, 1:2, function(p) rbinom(1, 1, p))
  A[lower.tri(A)] <- 0
  diag(A) <- 0
  A <- A + t(A)
  return(A)
}
```

@@

```{r}
K <- 2
p <- K * (K + 1) / 2
q <- K * (K - 1) / 2
alpha <- .5
a1 <- 2
b1 <- 1
a2 <- 1
b2 <- 2
```

@@

```{r}
n.vec <- c(64, 128, 256, 512, 1024)
iter <- 100
```

```{r}
draw.pabm.beta.2 <- function(n, a1, b1, a2, b2, alpha = .5) {
  z <- sample(c(1, 2), n, prob = c(alpha, 1 - alpha), replace = TRUE)
  z <- sort(z)

  X <- cbind(c(lambda11, rep(0, n2)),
             c(lambda12, rep(0, n2)),
             c(rep(0, n1), lambda21),
             c(rep(0, n1), lambda22))
  Y <- cbind(c(lambda11, rep(0, n2)),
             c(rep(0, n1), lambda21),
             c(lambda12, rep(0, n2)),
             c(rep(0, n1), lambda22))
  P <- X %*% t(Y)
  A <- draw.graph(P)

  return(list(A = A, z = z))
}

normalized.laplacian <- function(W) {
  n <- nrow(W)
  D.neg.sqrt <- diag(colSums(W) ** -.5)
  I <- diag(n)
  return(I - D.neg.sqrt %*% W %*% D.neg.sqrt)
}

cluster.pabm <- function(A, K) {
  p <- K * (K + 1) / 2
  q <- K * (K - 1) / 2
  V <- eigen(A, symmetric = TRUE)$vectors[, c(seq(p), seq(n, n - q + 1))]
  V <- V / mean(V)
  B <- abs(V %*% t(V))
  # L <- graph.laplacian(B)
  L <- normalized.laplacian(B)
  eigenmap <- eigen(L, symmetric = TRUE)$vectors[, seq(n - 1, n - K)]
  clustering <- mclust::Mclust(eigenmap, K)$classification
}
```

```{r fig.cap = 'IQR clustering error using Algorithm 1 for sample sizes from 64 to 1024. Simulations were repeated 100 times for each sample size.', cache = TRUE}
set.seed(314159)

results.df <- foreach(n = n.vec,.combine = dplyr::bind_rows) %do% {
  plyr::ldply(seq(iter), function(i) {
    Az <- draw.pabm.beta.2(n, a1, b1, a2, b2, alpha)
    A <- Az$A
    z <- Az$z
    clustering <- cluster.pabm(A, K)
    error <- mean(clustering == z)
    if (error > .5) error <- 1 - error
    dplyr::tibble(n = n, error = error) %>% 
      return()
  }, .parallel = TRUE) %>% 
    return()
}

results.df %>% 
  dplyr::group_by(n) %>% 
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75)) %>% 
  ggplot() + 
  # scale_y_log10() + 
  labs(y = 'error') + 
  geom_line(aes(x = n, y = med.err)) + 
  geom_errorbar(aes(x = n, ymin = first.q, ymax = third.q))
```

@@

```{r, fig.cap = 'Between-cluster inner products of the eigenvectors of $A$ for varying sample sizes.', cache = TRUE}
densities.df <- plyr::ldply(n.vec, function(n) {
  Az <- draw.pabm.beta.2(n, a1, b1, a2, b2, alpha)
  A <- Az$A
  z <- Az$z
  n1 <- sum(z == 1)
  n2 <- sum(z == 2)
  V <- eigen(A, symmetric = TRUE)$vectors[, c(seq(3), n)]
  B <- V %*% t(V)
  inner.prods <- as.numeric(B[(n1+1):n, (n1+1):n])
  return(dplyr::tibble(n = n, inner.prods = inner.prods))
}, .parallel = TRUE)

@@

```{r}
rot.mat.2 <- function(angle, axis = 1) {
  # rotation matrix for K=2

  if (axis %in% c(1, 3)) {
    rotation <- matrix(c(cos(angle), -sin(angle), sin(angle), cos(angle)),
                       nrow = 2, ncol = 2, byrow = TRUE)
    if (axis == 1) {
      return(as.matrix(Matrix::bdiag(1, rotation, -1)))
    } else {
      return(as.matrix(Matrix::bdiag(rotation, 1, -1)))
    }
  } else if (axis == 2) {
    return(matrix(c(cos(angle), 0, sin(angle), 0,
                    0, 1, 0, 0,
                    -sin(angle), 0, cos(angle), 0,
                    0, 0, 0, -1),
                  nrow = 4, ncol = 4,
                  byrow = TRUE))
  } else {
    stop(simpleError('axis must be 1, 2, or 3'))
  }
}

hyp.rot.mat.2 <- function(angle, axis = 1) {
  # hyperbolic rotation matrix for $K=2

  if (axis == 1) {
    return(matrix(c(1, 0, 0, 0,
                    0, 1, 0, 0, 
                    0, 0, cosh(angle), sinh(angle),
                    0, 0, sinh(angle), cosh(angle)), 
                  byrow = TRUE,
                  nrow = 4, ncol = 4))
  } else if (axis == 2) {
    return(matrix(c(1, 0, 0, 0,
                    0, cosh(angle), 0, sinh(angle),
                    0, 0, 1, 0,
                    0, sinh(angle), 0, cosh(angle)),
                  byrow = TRUE,
                  nrow = 4, ncol = 4))
  } else if (axis == 3) {
    return(matrix(c(cosh(angle), 0, 0, sinh(angle),
                    0, 1, 0, 0, 
                    0, 0, 1, 0, 
                    sinh(angle), 0, 0, cosh(angle)),
                  byrow = TRUE,
                  nrow = 4, ncol = 4))
  } else {
    stop(simpleError('axis must be 1, 2, or 3'))
  }
}

construct.Q <- function(angles) {
  rot.angles <- angles[1:3]
  hyp.angles <- c(angles[1], angles[4], -angles[4])
  rot.matrices <- lapply(seq(3), function(i) rot.mat.2(rot.angles[i], i))
  hyp.matrices <- lapply(seq(3), function(i) hyp.rot.mat.2(hyp.angles[i], i))
  matrices <- c(rot.matrices, hyp.matrices)
  return(Reduce(`%*%`, matrices))
}

obj.fun <- function(angles, Z.hat, clusters) {
  R <- construct.Q(angles)

  XU.c <- Z.hat %*% R
  in.prods <- XU.c[clusters == 1, ] %*% t(XU.c[clusters == 2, , drop = FALSE])
  return(
    sum(XU.c[clusters == 1, 2] ** 2) + 
      sum(XU.c[clusters == 2, 1] ** 2) + 
      sum((XU.c[clusters == 1, 3] - XU.c[clusters == 1, 4]) ** 2) + 
      sum((XU.c[clusters == 2, 3] + XU.c[clusters == 2, 4]) ** 2)
  )
}

transform.embedding <- function(Z.hat, clusters, init.angles) {
  # init.angles <- runif(4, -pi, pi)
  # init.angles <- rep(0, 4)
  angles.1 <- optim(init.angles, function(x) obj.fun(x, Z.hat, clusters), 
                    method = 'SANN')$par
  angles.2 <- optim(angles.1, function(x) obj.fun(x, Z.hat, clusters),
                    method = 'Nelder-Mead',
                    control = list(abstol = 1e-10, maxit = 1e3))$par
  R <- construct.Q(angles.2)
  return(Z.hat %*% R)
}
```

@@

```{r lambda_est, cache = TRUE, fig.cap = 'Median and IQR RMSE from Algorithm 4 using sample sizes from 64 to 1024. Simulations were repeated 100 times for each sample size.'}
set.seed(314159)

Ipq <- diag(c(1, 1, 1, -1))

rmse.df <- foreach(n = n.vec, .combine = dplyr::bind_rows) %do% {
  plyr::ldply(seq(iter), function(i) {
    z <- sample(c(1, 2), n, prob = c(alpha, 1 - alpha), replace = TRUE)
    z <- sort(z)

    X <- cbind(c(lambda11, rep(0, n2)),
               c(lambda12, rep(0, n2)),
               c(rep(0, n1), lambda21),
               c(rep(0, n1), lambda22))
    Y <- cbind(c(lambda11, rep(0, n2)),
               c(rep(0, n1), lambda21),
               c(lambda12, rep(0, n2)),
               c(rep(0, n1), lambda22))
    P <- X %*% t(Y)
    A <- draw.graph(P)
    lambda.matrix <- cbind(c(lambda11, lambda21),
                           c(lambda12, lambda22))
    Zhat <- embedding(A, p, q)
    Phat <- Zhat %*% Ipq %*% t(Zhat)
    rmse <- lambda.rmse(lambda.matrix, Phat, z)
    dplyr::tibble(n = n, rmse = rmse) %>% 
      return()
  }, .parallel = TRUE) %>% 
    return()
}

>>>>

```{r}
import::from(magrittr, `%>%`)
library(plot.matrix)
library(ggplot2)
```

@@

@@

```{r fig.cap = 'IQR clustering error using Algorithm 1 for sample sizes from 64 to 1024. Simulations were repeated 100 times for each sample size.', cache = FALSE}
clustering.df <- readr::read_csv('clustering.csv')
clustering.df %>%
  dplyr::group_by(n) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75)) %>%
  ggplot() +
  # scale_y_log10() +
  labs(y = 'error') +
  geom_line(aes(x = n, y = med.err)) +
  geom_errorbar(aes(x = n, ymin = first.q, ymax = third.q))
```

@@

```{r, fig.cap = 'Between-cluster inner products of the eigenvectors of $A$ for varying sample sizes.', cache = FALSE}
densities.df <- readr::read_csv('densities.csv')

@@

@@

```{r lambda_est, cache = TRUE, fig.cap = 'Median and IQR RMSE from Algorithm 4 using sample sizes from 64 to 1024. Simulations were repeated 100 times for each sample size.'}
rmse.df <- readr::read_csv('rmse.csv')
